Notes 2.0 on presentation of model for co-evolution

### 2017-11-26 code-LOT-extension/co-evo/co-evolution-notes2.tex
update, filled quiver plots for co-evo illustration

\title{Notes on presentation of model for co-evolution}
\author{Thomas Brochhagen}
\date{}

@@

\subsection*{Learning bias}
In the paper, the learners' inductive bias is a function of the complexity of the formulae that types lexicalize. Introducing our measure and applying it only to this small type space is probably not the way to go. For illustratory purposes, we may instead assume that there's an abstract parameter (which we later make concrete using our measure) $c \geq 1$ that regulates how strongly $\Llack$ is preferred over $\Lbound$. In what follows, I assumed this to be $P(L_i) \propto c$ if $L_i$ is $\Llack$ and otherwise $1$. For instance, if we have  a type space with $\tau_i$ as literal $\Lbound$ and $\tau_j$ as pragmatic $\Llack$, then $P(\tau_i) \approx .33$ for $c = 2$. This is just for illustration, of course. We can change this at will. As far as I can see, the reduced type spaced behave in a very predictable manner. I set $c = 2$ for all the plots below.

\subsection*{General remarks}
First, I did not succeed in putting the model in an elegant closed form. The good news is that the type space is small enough so that we don't need to sample anything. In particular, we don't need to sample from speaker probabilities to approximate $Q$ for large $k$; we can look at all possible state-message sequences instead, meaning that we get the same result each time we apply replicator or mutator steps.

Second, there are some issues with phase-portrait-like representations. The main problem is that I have not (yet?) found an algorithm that succeeds in reliably finding stable points for mutator (or replicator-mutator) steps.

Third, one of the major disadvantages of the smaller type spaces is that (I fear that) they give the impression that the mutator-step is doing most the work. As shown below, $M(x)$ is very close to $M(RD(x))$. This is not true of the bigger type space, but makes sense in these reduced ones. After all, learning is what differentiates $\Llack$ from $\Lbound$. I'm concerned that these results may give a wrong impression.

Lastly, I think that all the representations I generated have some negative sides (details below). Before improving on them, I thought we should discuss these matters, making use of the fact that their flaws are evident.

\subsection*{Phase portraits}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{phases-lam01-bias2-l01-k5}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{phases-lam20-bias2-l10-k5}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Approximate phase-portraits for (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$}
\label{fig:phase}
\end{figure}

Figure \ref{fig:phase} shows $8$ (approximate) portraits for applications of replicator steps only, mutator steps only, and replicator-mutator steps on our four type-pairings, for two parameter constellations. For instance, the upper-left-most portrait shows the dynamics for a type space consisting of literal $\Lbound$ ($x$) and literal $\Llack$ ($1-x$). We can see that, as long as there are some $\Lbound$ types in the population, the population progresses toward the extreme where only this type survives.

I used a couple of heuristics to generate these plots. First, I took $1000$ points in $[0,1]$ and applied the dynamic in question. Then, I checked if there were any points for which $D(x) = x$. As expected, the only stable points found in this way are the extremes for the replicator step: $RD(1) = 1$ and $RD(0) = 0$. To check whether these points are attractors, I looked at adjacent points to figure out whether they moved toward or away from that point. I did the same for the arrows, evenly spread across the axis.
\begin{itemize}
\item[{\bf The good}] It's easy to see how the dynamics play out. 
\item[{\bf The bad}] It's not easy to find stable points when mutation is involved. As mentioned above, I tried a couple of optimization algorithms but didn't get anywhere so far. Additionally, $M(x)$ and $M(RD(x))$ are (almost) indistinguishable. This is obviously an issue for all the depictions I played around with. It's not as obvious in these ones though.
\item[{\bf The ugly}] There are {\em a lot} of points where $RD(x) = x$ in the type space consisting of only literal and pragmatic $\Lbound$. This explains why these two phase portraits look so ugly; there are many overlapping circles.
\end{itemize}
{\bf My take:} I think these plots are nice but that they require the most cleaning. The major advantage I see here is that, even without finding stable points, it's easy to see where $M(\cdot)$ and $M(RD(\cdot))$ are headed (having less arrows would make this a lot easier to see). This representation slightly masks my big concern of mutator- and RM-steps being almost the same. As for the problem of too many stable points: We could be a little more selective about which points to depict (e.g., relative to the basin of attraction) or use other pictorial means (see the last set of plots).

\subsection*{Simple plots}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{plot-lam01-bias2-l01-k5}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{plot-lam20-bias2-l10-k5}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$. The $x$-axis shows proportion of $x$ before the update, the $y$-axis shows the result of a step of the dynamic on this $x$-value.}
\label{fig:plots}
\end{figure}

Figure \ref{fig:plots} shows $8$ plots using the same data as above. The main difference is that there's no information about stable points and that the directionality of change, particularly in the case of replication only, is harder to see.

\begin{itemize}
\item[{\bf The good}] It's very easy to see the effects of mutation only and the RMD. In particular, that (i) all of $Q$'s cells are equal to $.5$ in populations of only literal and pragmatic $\Lbound$, (ii) but that there is a clear difference in the learnability of literal and pragmatic $\Llack$ due to differences in their likelihoods when $\lambda > 1$.
\item[{\bf The bad}] It's not evident that there are differences across applications of the replicator step. Additionally, these plots may give the impression that replication {\em in general} is doing very little. Particularly when it comes to a comparison between $M(x)$ and $RMD(x)$. I must concede that it is true that replication is doing very little in these reduced type spaces.
\end{itemize}

{\bf My take:} I don't think that we should use plots like these as they don't serve the illustrative purpose we're after. I included them because they clarified a lot of questions I had when I first looked at the phase portraits.

\subsection*{Dynamics}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{quiver-lam1-c2-k5-l1}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{quiver-lam20-c2-k5-l10}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$. Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
\label{fig:quiver}
\end{figure}

Lastly, I wanted to try my hand at compressing the information more while trying to highlight the main take away of the dynamics applied to these reduced type spaces. Figure \ref{fig:quiver} is an illustration. I used the sketch Michael sent us with his comments as inspiration. We think of the type space as being $2$ dimensional. Along the four edges we have same types competing as we had in the previous plots. This $2D$ space is ambiguous once you move from the edges, so I only applied the dynamics on edges. More precisely, in each plot, the $x$-coordinate indicates the proportion of literal types and the $y$-coordinate indicates the proportion of $\Llack$ users. Points that are not on edges are ambiguous because, e.g., $(.5,.5)$ could mean $50\%$ are literal $\Lbound$ and $50\%$ are pragmatic $\Llack$ but also that there are $25\%$ of each of the four types in the population. The edges are unambiguous because each population consists of maximally two types, opposed with respect to a single dimension. This explains why replication in Figure \ref{fig:quiver} only moves along edges.

\begin{itemize}
\item[{\bf The good}] These plots look very similar to what Michael sketched out by hand. The RD plots are also a lot cleaner than the phase portraits I made above.
\item[{\bf The bad}] The ambiguity of $(x,y)$ away from edges.
\item[{\bf The ugly}] Its painfully evident that mutation steps and RMD steps are almost identical. Additionally, vectors are not scaled. Right now, they are automatically resized relative to the other vectors on their edge. This makes it easier to show small effects. If I scale the vectors then the RD plots look as if they consisted only of points because the mutator-steps are much larger. Note also that the movement from pragmatic $\Lbound$ to literal $\Lbound$ under replication only is only apparent when we look at enough points along the edge. However, this makes the plots look cluttered everywhere else. This is fixable and mostly cosmetic though.
\end{itemize}
{\bf My take:} I like these the most. Of course, we could also generate each edge separately and present them in a format similar to that of phase portraits above.

I couldn't come up with anything, but: Is there a space in which we could disentangle the ambiguity of the $2D$ space? That is, a shape that is as the regular $3$-simplex is to the regular $2$-simplex?

>>>>

\title{Notes 2.0 on presentation of model for co-evolution}
\author{Thomas Brochhagen}
\date{}

@@

\subsection*{Learning bias}
In the paper, the learners' inductive bias is a function of the complexity of the formulae that types lexicalize. Introducing our measure and applying it only to this small type space is probably not the way to go. For illustratory purposes, we may instead assume that there's an abstract parameter (which we later make concrete using our measure) $c \geq 1$ that regulates how strongly $\Llack$ is preferred over $\Lbound$. In what follows, I assumed this to be $P(L_i) \propto c$ if $L_i$ is $\Llack$ and otherwise $1$. For instance, if we have  a type space with $\tau_i$ as literal $\Lbound$ and $\tau_j$ as pragmatic $\Llack$, then $P(\tau_i) \approx .33$ for $c = 2$. This is just for illustration, of course. We can change this at will. As far as I can see, the reduced type spaced behaves in a very predictable manner.% I set $c = 2$ for all the plots below.

\subsection*{General remarks 2.0}
There are three remarks to be made about this update.

First, the phase plots are a lot cleaner and all show stationary points. These were either calculated, or estimated as being in the middle of two points that changed in opposite directions. As before, the RD competition between literal and pragmatic $\Lbound$ is messy because there are many points for which $RD(x) = x$. On this front, I just changed white circles with black edges to red circles without edges for better visibility when many circles overlap.

Second, when it comes to the direct competition of only two types on a line (phase plots), M and RMD still produce pretty similar outcomes. I tested a range of learning bias values and larger $k$ but there's not much change. This is not really troublesome (it being a consequence of there only being two types in a population), but I think this may disqualify the phase plots as candidates for a good illustration of the model's dynamics.

Third, I tried my hand at the 2D quiver plots with all types, not only at the edges but also at the center of the unit square. As Michael suggested, $(x,y)$-pairs in the square give the following proportions of types:
\begin{itemize} 
  \item $lit. \; \Lbound$: $(1-x) * (1-y)$ 
  \item $lit. \; \Llack$: $(1-x) * y$
  \item $prag.\; \Lbound$: $x * (1-y)$
  \item $prag.\; \Llack$: $x * y$
\end{itemize}

Overall, I think these plots look good. Important, to me, is that (i) M and RMD are clearly distinguishable and that (ii) it's clear that RD with lowish $\lambda$ leads to $\Lbound$. Another plus is that it's easy to explain how this plot came about. On the negative side, I'm still a bit unhappy about having to assume that prag. and $\Llack$ are independent, but this is clearly better than just plotting the edges. The one important I am missing from these plots is clear movement on the lower edge, from $(1,0)$ to $(0,0)$, which would indicate movement from prag. $\Lbound$ to lit. $\Lbound$,  at least for RD-only. This would help us in driving the message home that pragmatic recruitment is not functionally advantageous for all types. I think the problem right now is that movement on this edge progresses in very small steps (if at all; cf. phase plots). These small steps disappear once we scale them with the bigger change that happens elsewhere in the square. Reinterpreting $(x,y)$ would not help on the edge itself, but maybe taking average trajectories would at least make points close to the edge move toward lit. $\Llack$ rather than toward its pragmatic counterpart.

\subsection*{Phase portraits}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.3]{phases-lam01-bias105-l01-k8}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.3]{phases-lam20-bias105-l10-k8}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Approximate phase-portraits for (a) $\lambda = 1, c = 1.05, l = 1, k = 8$ and (b) $\lambda = 20, c = 1.05, l =10, k =8$. Attractors are in black. Non-attractor points are in red for better visibility for lit. $\Lbound$ vs prag. $\Lbound$, which has many.}
\label{fig:phase}
\end{figure}

As before, Figure \ref{fig:phase} shows $8$ (approximate) portraits for applications of replicator steps only, mutator steps only, and replicator-mutator steps on our four type-pairings, for two parameter constellations ($c$ is now $1.05$ instead of $2$). For instance, the upper-left-most portrait shows the dynamic for a type space consisting of literal $\Lbound$ ($x$) and literal $\Llack$ ($1-x$). We can see that, as long as there are some $\Lbound$ types in the population, the population progresses toward the extreme where only this type survives.

Beyond having cleaner portraits, my take on these is as before: They are nice because it's easy to see how the dynamics play out but they do not highlight difference between $M(x)$ and $M(RD(x))$ very well. The most pronounced differences I have found are when $\lambda$ is high, $c$ low, and learners tend to sample from the posterior posterior sampling (e.g.; $\lambda = 20$, $c = 1.05$, $l = 1$). However, that's a rather esoteric parameter constellation to illustrate the dynamics with.

\subsection*{Dynamics}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{quiver-lam1-c105-k8-l1}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{quiver-lam20-c105-k8-l10}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 1.05, l = 1, k = 8$ and (b) $\lambda = 20, c = 1.05, l =10, k =8$. Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
\label{fig:quivera}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{quiver-lam1-c200-k8-l1}
  \caption{}
  \label{fig:sub1b}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{quiver-lam20-c200-k8-l10}
  \caption{}
  \label{fig:sub2b}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 8$ and (b) $\lambda = 20, c = 2, l =10, k =8$. Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
\label{fig:quiverb}
\end{figure}

Figure \ref{fig:quivera} ($c = 1.05$) and Figure \ref{fig:quiverb} ($c = 2$) illustrate the new filled unit squares. The main news are that we got rid of the ambiguity for $(x,y)$ when we move away from the edges (see above). Taking stock: the good is that (i) the contrast between RD, M, and RMD is clear, (ii) we can see that lit. $\Llack$ is functionally disadvantageous and that its pragmatic counterpart is so as well if $\lambda$ is low, (iii) that a functional advantage {\em and} a learnability advantage will select for pragmatic $\Llack$ but not one of these on its own. The bad is that we don't see the slight functional disadvantage that prag. $\Lbound$ suffers from when compared to its literal counterpart.
